---
title: ''
csl: mbio.csl #Get themes at https://github.com/citation-style-language/styles
fontsize: 11pt
output:
  pdf_document:
    includes:
      in_header: header.tex
    keep_tex: yes
  word_document: default
  html_document:
    df_print: paged
geometry: margin=1.0in
bibliography: references.bib
---


```{r knitr_settings, eval=TRUE, echo=FALSE, cache=FALSE, warning=FALSE, message=FALSE, tidy=TRUE}
options(tidyverse.quiet = TRUE)

library(tidyverse)
library(knitr)
library(here)

opts_chunk$set("tidy" = TRUE)
opts_chunk$set("echo" = FALSE)
opts_chunk$set("eval" = TRUE)
opts_chunk$set("warning" = FALSE)
opts_chunk$set("message" = FALSE)
opts_chunk$set("cache" = FALSE)

inline_hook <- function(x){
	print(x)

	if(is.list(x)){
		x <- unlist(x)
	}

	if(is.numeric(x)){
		if(abs(x - round(x)) < .Machine$double.eps^0.5){
			paste(format(x,big.mark=',', digits=0, scientific=FALSE))
		} else {
			paste(format(x,big.mark=',', digits=2, nsmall=2, scientific=FALSE))
		}
	} else {
    	paste(x)      
	}
}
knitr::knit_hooks$set(inline=inline_hook)
```



\vspace{35mm}

# An osmotic laxative renders mice susceptible to prolonged *Clostridioides difficile* colonization and hinders clearance 

\vspace{35mm}


Sarah Tomkovich^1^, Ana Taylor^1^, Jacob King^1^, Joanna Colovas^1^, Lucas Bishop^1^, Kathryn McBride^1^, Sonya Royzenblat^1^, Nicholas A. Lesniak^1^, Ingrid L. Bergin^2^, Patrick D. Schloss^1$\dagger$^

\vspace{40mm}

$\dagger$ To whom correspondence should be addressed: pschloss@umich.edu

1\. Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI, USA

2\. The Unit for Laboratory Animal Medicine, University of Michigan, Ann Arbor, MI, USA


\newpage
\linenumbers


## Abstract
(Modify depending on target journal, currently abstract submitted to World Microbe Forum) Antibiotics are a major risk factor for *Clostridioides difficile* infections (CDIs) because of their impact on the intestinal microbiome. However, non-antibiotic medications such as the ubiquitous osmotic laxative polyethylene glycol (PEG) 3350, also alter the microbiota, but whether PEG impacts CDI susceptibility and clearance is unclear. To examine how PEG impacts susceptibility, we treated C57Bl/6 mice with 5-day and 1-day doses of 15% PEG in the drinking water and then challenged the mice with *C. difficile* 630 spores. We used clindamycin-treated mice as a control because they consistently clear *C. difficile* within 10 days post-infection (dpi). To examine how PEG treatment impacts clearance, we administered PEG for 1 day to clindamycin-treated, *C. difficile*-challenged mice either immediately following challenge or 3 dpi. We collected longitudinal stool samples to examine *C. difficile* levels in the stool via anaerobic culture and profiled the microbiota by 16S rRNA sequencing. PEG treatment alone was sufficient to render mice susceptible to CDI and 5-day PEG-treated mice remain colonized for up to 30 dpi. Additionally, 5-day PEG treated mice remained susceptible to CDI 10-days post treatment. In contrast, 1-day PEG treated mice were transiently colonized, clearing *C. difficile* within 7 dpi. Although 5-day PEG-treated mice exhibited prolonged *C. difficile* colonization, we saw no difference in histological inflammation between PEG- and clindamycin-treated mice. Additionally, administering PEG to mice after *C. difficile* challenge prolonged colonization up to 30 dpi in mice that received PEG immediately after challenge and 15 dpi in mice that received PEG 3 dpi. When we examined microbiota composition across our different treatment groups, we found decreased richness in the PEG-treated mice that exhibited prolonged *C. difficile* colonization. Importantly, there were increased Bacteroides and Enterobacteriaceae and decreased Lachnospiraceae and Oscillibacter in most of the PEG-treated mice with prolonged *C. difficile* colonization. Our findings suggest the osmotic laxative PEG 3350 alters the mouse microbiota and disrupts colonization resistance to *C. difficile*, as well as clearance in mice with a CDI. Considering that most hospitals recommend not performing *C. difficile* testing on patients taking laxatives and laxatives are used when administering fecal microbiota transplants via colonoscopy to patients with recurrent CDIs, further studies are needed to evaluate if laxatives impact human microbiota colonization resistance.

\newpage

## Introduction

Antibiotics are a major risk factor for *Clostridioides difficile* infections (CDIs) because they disrupt microbiota colonization resistance [@Britton2014]. However, antibiotics are not the only types of medications that disrupt the microbiota [@Maier2018; @LeBastard2017; @VichVila2020]. Although, other medications (proton pump inhibitors, osmotic laxatives, antimotility agents, and opioids) have been implicated as risk or protective factors for CDIs through epidemiological studies, whether the association is due to their impact on the microbiome is still unclear [@Oh2018; @Mora2012; @Nehra2018; @Krishna2013; @Tomkovich2019]. 

Many of the non-antibiotic medications associated with CDIs are known to modulate gastrointestinal motility leading to either increased or decreased colonic transit time, which in turn also strongly impacts microbiota composition and function [@Vandeputte2015; @VujkovicCvijin2020]. Stool consistency often serves as an approximation of intestinal motility. Our group has shown that when *C. difficile* negative controls are separated into two groups based on stool consistency, there are shared microbiota features such as lower alpha diversity in samples from CDI patients and control patients with diarrhea compared to control samples that were *C. difficile* negative with non-diarrheal consistency [@Schubert2015]. These results led to a hypothesis that bacterial communities from patients experiencing diarrhea are susceptible to developing CDIs.

Osmotic laxatives can lead to diarrhea depending on the administered dose and temporarily disrupt the human intestinal microbiota [@Nagata2019]. The ubiquitous osmotic laxative, polyethylene glycol (PEG) 3350 is found in Miralax, Nulytely, and Golytely and is also commonly used as bowel preparation for colonoscopies. Interestingly, previous studies have shown that treating mice with PEG alone altered microbiota composition, reduced acetate and butyrate production, altered the mucus barrier, and rendered the mice susceptible to *C. difficile* infection [@Kashyap2013; @Ferreyra2014; @Tropini2018; @VanInsberghe2020]. The mucus barrier is thought to mediate protection from *C. difficile* infections by protecting intestinal epithelial cells from the toxins produced by *C. difficile* (Ref). However, whether laxative results in more severe CDIs in mice and how long mice remain colonized with *C. difficile* after challenge is unclear.

Beyond susceptibility, PEG is also relevant in the context of treating recurrent CDIs via fecal microbiota transplant (FMT) where a healthy microbiota is administered to the patient to restore colonization. For FMTs that are delivered via colonoscopy, patients typically undergo bowel preparation by taking an osmotic laxative prior to the procedure. Many of the FMT studies to date rationalize the use of laxatives (Ref) based on a 1996 case study with 2 pediatric patients where the authors suggested in the discussion that the laxative may help flush *C. difficile* spores and toxins from the intestine [@Liacouras1996].

In the past, our group has used C57BL6 mice to characterize how antibiotics including clindamycin disrupt the microbiota and influence *C. difficile* susceptibility and clearance [ref]. Although, two groups have now shown PEG treatment alone renders mice susceptible to *C. difficile*, these studies have raised additional questions regarding the dynamics and severity of infection as well as the role of laxative treatment in *C. difficile* clearance that should be addressed to better inform how we think about laxatives in the context of CDIs. Here, we used our C57BL/6 clindamycin model as a control group to characterize how long PEG-treated mice remain susceptible, whether PEG treatment results in more severe CDI and sustained *C. difficile* colonization, and if PEG treatment post-CDI can promote *C. difficile* clearance.

## Results

**5-day laxative treatment leads to prolonged *C. difficile* colonization in mice.** We compared PEG-treated mice to our standard 10 mg/kg clindamycin treatment, which temporarily renders the mice susceptible to *C. difficile*, with mice typically clearing *C. difficile* within 10 days post-infection [@Tomkovich2019; @Tomkovich2020]. All PEG-treated mice were administered a 15% PEG solution in the drinking water for 5-days, one group was also treated with clindamycin, and one group was allowed to recover for 10 days prior to challenge (Fig. 1A). PEG treatment resulted in weight loss in all 3 groups of PEG-treated mice, with the greatest change in weight observed on the fifth day of PEG treatment (Fig. 1B). After either PEG, clindamycin, or PEG and clindamycin treatment all mice were challenged with 10^3^ *C. difficile* 630 spores. All treatments rendered mice susceptible to *C. difficile* colonization (Fig. 1C), however PEG-treated mice remained colonized at a high level through 30 days post-infection. In contrast, the clindamycin-treated mice that cleared *C. difficile* within 10 days post-infection. Surprisingly, mice were still susceptible to *C. difficile* infection after 10-days of recovery from PEG treatment although *C. difficile* was not detectable in most of the group in the initial 5 days post-infection (Fig. 1C). From 9 days post-infection onward, the median *C. difficile* stabilized for the 5-day PEG plus 10-day recovery group of mice and remained high through 30 days post-infection (Fig. 1C). Thus, osmotic laxative treatment alone was sufficient to render mice susceptible to prolonged *C. difficile* colonization and PEG-treated mice remained susceptible for up to 10 days post PEG treatment.
**5-day laxative treatment differentially disrupts the fecal microbiota compared to clindamycin treatment.** Given 
**5-day laxative treatment does not promote more severe CDIs despite altering the mucosal microbiota.** Given 

* Figure 1. 5-day PEG treatment prolongs susceptibility and mice become persistently colonized with *C. difficile*.

* Figure 2. 5-day PEG treatment disrupts the stool microbiota for a longer amount of time compared to clindamycin-treated mice.

* Figure S1. 5-day PEG treatment plus 10-day recovery mice microbiota dynamics post-infection.

* Figure 3. 5-day PEG treatment does not result in more severe CDIs, although mucosal microbiota is altered. 

* Figure 4. 1-day PEG treatment renders mice susceptible to transient *C. difficile* colonization.

* Figure 5. 1-day PEG treatment post *C. difficile* challenge prolongs colonization regardless of whether an FMT is also administered.

* Figure 6. Specific microbiota features associated with prolonged *C. difficile* colonization in PEG treated mice.

* Figure S2. Specific OTUs associated with clearance that are mostly absent in mice with prolonged *C. difficile* colonization.
  + Ex. *Muribaculum intestinale*. 

* Figure 7. Schematic summarizing findings.

## Discussion

* Summary of major findings

* Discussion of prolonged persistence. C. difficile sequences detected in tissue samples. Association with mucin-degrading bacteria suggested by recent papers.

* Discuss why we might not have observed more severe histology in PEG mice relative to clindamycin-treated mice
  + Antibiotics may also impact mucus layer
  + Strain of bacteria used

* Protective bacteria missing in PEG-treated mice

* Discuss what these findings might mean for human patients
  + What's known regarding laxatives and susceptibility to CDIs
  + Relevance to human FMTs? Unclear what the best administration route is because there have been no studies designed to evaluate the best administration route for FMTs.

## Conclusions

## Acknowledgements
We thank members of the Schloss lab for feedback on planning the experiments and data presentation. We also thank Andrew Henry for help with media preparation and bacterial culture. We also thank the Unit for Laboratory Animal Medicine at the University of Michigan for maintaining our mouse colony and providing the institutional support for our mouse experiments. Finally, we thank Kwi Kim, Austin Campbell, and Kimberly Vendrov for their help in maintaining the Schloss lab's anaerobic chamber. This work was supported by the National Institutes of Health (U01AI124255). ST was supported by the Michigan Institute for Clincial and Health Research Postdoctoral Translation Scholars Program (UL1TR002240 from the National Center for Advancing Translational Sciences). 

## Materials and Methods


\newpage
```{r sample_size for figure legends}


test <- 4


```

\includegraphics{figure_1.pdf}

**Figure 1. 5-day PEG treatment prolongs susceptibility and mice become persistently colonized with *C. difficile*.** A. Setup of the experimental timeline for subset of experiments with 5-day PEG treated mice. B. Weight change from baseline weight in groups after treatment with PEG and/or clindamycin, followed by *C. difficile* challenge. C. *C. difficile* CFU/gram stool measured over time (N = `r test`-`(insert variable name)` mice per timepoint) via serial dilutions. The black line represents the limit of detection for the first serial dilution. CFU quantification data was not available for each mouse due to stool sampling difficulties (particularly the day the mice came off of the PEG treatment) or early deaths. Lines represent the median for each source and circles represent individual mouse samples. Asterisks indicate timepoints where the weight change or CFU/g was significantly different between groups by the Kruskal-Wallis test with Benjamini-Hochberg correction for testing multiple timepoints.
\newpage

\includegraphics{figure_2.pdf}
**Figure 2. 5-day PEG treatment disrupts the stool microbiota for a longer amount of time compared to clindamycin-treated mice.** A. 
\newpage

\includegraphics{figure_3.pdf}
**Figure 3. 5-day PEG treatment does not result in more severe CDIs, although mucosal microbiota is altered.** A. 
\newpage

``` {r figure_4_permnova}
one_day_permnova_results <- read_tsv(here("data/process/1_day_PEG_permanova.tsv"))
one_day_top_2_contrib <- one_day_permnova_results %>% arrange(desc(R2)) %>%
  top_n(3, R2)
one_day_top_contrib_name = (one_day_top_2_contrib[2,1])
one_day_sec_contrib_name = (one_day_top_2_contrib[3,1])
top_R2 = one_day_top_2_contrib[2,6]
sec_R2 = one_day_top_2_contrib[3,6]
```
\includegraphics{figure_4.pdf}
**Figure 4. 1-day PEG treatment renders mice susceptible to transient *C. difficile* colonization.** A. Setup of the experimental timeline for the 1-day PEG treated subset of mice. B. CFU/gram stool measured over time (N = 6 mice per timepoint) via several dilutions. The black dotted line represents the limit of detection for the first serial dilution. Asterisks indicate timepoints where the CFU/gram was siginificantly different between groups using the Kruskall-Wallis test with a Benjamini-Hochberg correction for multiple timepoints. C. Principle Coordinate Analysis plot of the groups over time with the alpha representating the same time scale as in panel D (`r one_day_top_contrib_name`: R^2^ = `r top_R2`; `r one_day_sec_contrib_name`: R^2^ = `r sec_R2`). D. Shannon Diveristy Index of the groups over time. Only days with samples from all groups are shown. Samples for some mice were difficult to obtain 
due to the laxative treatment. The alpha scale follows accordingly with the timeline. E. Line plots of relative percent abundance of selected genera over time. Only days with samples from all groups shown. The gray line represents the limit of detection. 

\includegraphics{figure_5.pdf}
\newpage
\includegraphics{figure_5_16S.pdf}
**Figure 5. 1-day PEG treatment post C. difficile challenge prolongs colonization regardless of whether an FMT is also administered.** A. 
\newpage

\includegraphics{figure_6.pdf}
**Figure 6. Specific microbiota features associated with prolonged *C. difficile* colonization in PEG treated mice.** A. 
\newpage

\includegraphics{figure_7.pdf}
**Figure 7. Schematic summarizing findings.** A. 
\newpage

\includegraphics{figure_S1.pdf}
**Figure S1. 5-day PEG treatment plus 10-day recovery mice microbiota dynamics post-infection.** A. 
\newpage

\includegraphics{figure_S2.pdf}
**Figure S2. Specific OTUs associated with clearance that are mostly absent in mice with prolonged *C. difficile* colonization. Ex. *Muribaculum intestinale*.** A. 
\newpage

## References
